On Adding Pattern Matching to Haskell-based Deeply Embedded Domain Specific Languages

### 2020-10-08 paper/paper.tex
Line numbers
update latex
Reformat some example code, comment-out appendixes, move Conclusion to after Related Work

%% This is file `sample-acmsmall.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `acmsmall')
%% 
%% IMPORTANT NOTICE:
%% 
%% For the copyright see the source file.
%% 
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-acmsmall.tex.
%% 
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%% 
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%% The first command in your LaTeX source must be the \documentclass command.
\RequirePackage{amsmath,amssymb}
% \RequirePackage{stmaryrd}
\documentclass[runningheads, a4paper]{llncs}
% \usepackage{biblatex}
% \usepackage[switch]{lineno}
\usepackage{booktabs}
\usepackage{listings}
\usepackage{tcolorbox}
\usepackage{tikz}
\usepackage{tkz-graph}
\usepackage{stmaryrd}
% \usepackage{amsmath}

@@

% \renewcommand\linenumberfont{\normalfont\tiny\sffamily\color{red}}

@@

\newcommand{\showtodos}{}  % Comment out to hide todos

@@

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
% Say what problem is
  Capturing control flow is the Achilles heel of Haskell-based
  deeply embedded domain specific languages.
  Rather than use
  the builtin control flow mechanisms, artificial control flow combinators
  are used instead.
% Say why it is an interesting problem
  However, capturing traditional control flow in an deeply embedded domain specific language
  would support the writing of programs in a natural style by allowing the programmer to use the
  constructs that are already builtin to the base language, such as pattern
  matching and recursion.
% Say what our solution achieves
  In this paper, we expand the capabilities of
  Haskell-based deep embeddings with a compiler extension
  for reifying conditionals and pattern matching.
% Here is why the world will be a better place
  With this new support, the Haskell that we capture for expressing
  deeply embedded domain specific languages can be cleaner, Haskell-idiomatic,
  and more declarative in nature.

@@

\begin{lstlisting}[deletekeywords={Ord}]
import Data.Char (ord)

x :: Either Char Int
x = Left 'a'

example :: Int
example =
  externalize
    (case x of
      Left  c -> ord c
      Right i -> i)

@@

\begin{todo}
Should we have a diagram somewhere in the paper to show how these pieces all fit together?
\end{todo}

@@

% \section{Citations and Bibliographies}

@@

\appendix
\section{Additional Code for Standard Haskell Semantics Implementation}
\begin{lstlisting}[basicstyle=\tiny]
absEnv :: forall t. Env -> E t -> t
absEnv env (CaseExp x f) =
  sumMatchAbs f (absEnv env x)

absEnv env (TailRec f) = \ x ->
  case absEnv env f x of
    Step x' -> absEnv env (TailRec f) x'
    Done r  -> r

absEnv env (Var v) =
  case envLookup env v of
    Just x -> absEnv env
    Nothing ->
      error
        ("No binding for name "
          ++ show v)

absEnv env (Lam (name :: Name a)
                (body :: E b)) =
  \ (arg :: a) ->
    let go :: forall x. E x -> E x
        go expr@(Var name2 :: E a') =
          case namesEq name name2 of
            Just Refl -> rep arg
            _ ->
              case envLookup env name2 of
                Just v  -> v
                Nothing -> expr

        -- [... traverse rest of expression
        --      in go ...]
    in
    absEnv (extendEnv env
                      (name :=> rep arg))
           (go body)

absEnv env (LeftExp x) =
  Left (absEnv env x)

absEnv env (RightExp y) =
  Right (absEnv env y)

absEnv env (PairExp x y) =
  (absEnv env x, absEnv env y)

absEnv env (ConstructRep x) =
  unrep' (absEnv env x)

...
\end{lstlisting}

% \vfill\break

\section{A small EDSL for C code generation used internally by the C backend}
\begin{lstlisting}[basicstyle=\tiny]
type CCode = String
type CName = String

stmt :: CCode -> CCode
stmt = (<> ";")

cCall :: CName -> [CCode] -> CCode
cCall fnName args =
  fnName
  <> "(" <> intercalate ", " args <> ")"

derefPtr :: CCode -> CCode
derefPtr x = "*(" <> x <> ")"

cCast :: CCode -> CCode -> CCode
cCast toType x =
  "((" <> toType <> ")" <> x <> ")"

cIndex :: CCode -> Int -> CCode
cIndex x i =
  "(" <> x <> ")[" <> show i <> "]"

(!) :: CCode -> Int -> CCode
(!) = cIndex

(=:) :: CCode -> CCode -> CCode
x =: y = stmt (x <> " = " <> y)

cDecl :: CName -> CName -> CCode
cDecl cType v = stmt (cType <> " " <> v)

(#) :: CCode -> CName -> CCode
x # fieldName = x <> "." <> fieldName

macroDef
  :: CName -> [CName] -> [CCode]
     -> CCode
macroDef name args theLines =
  unlines
    [ "#define " <> name
              <> "(" <>
                intercalate ", " args
              <> ")\\"
    , "  do {\\"
    , unlines
        . map ("    "++)
        . map (++"\\")
        $ theLines
    , "  } while (0);"
    ]
\end{lstlisting}
% \begin{todo}
%   Fill in
% \end{todo}

>>>>

%% This is file `sample-acmsmall.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `acmsmall')
%% 
%% IMPORTANT NOTICE:
%% 
%% For the copyright see the source file.
%% 
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-acmsmall.tex.
%% 
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%% 
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%% The first command in your LaTeX source must be the \documentclass command.
\RequirePackage{amsmath,amssymb}
% \RequirePackage{stmaryrd}
\documentclass[runningheads, a4paper]{llncs}
% \usepackage{biblatex}
\usepackage{lineno}
\usepackage{booktabs}
\usepackage{listings}
\usepackage{tcolorbox}
\usepackage{tikz}
\usepackage{tkz-graph}
\usepackage{stmaryrd}
% \usepackage{amsmath}

@@

\renewcommand\linenumberfont{\normalfont\tiny\sffamily\color{red}}

@@

% \newcommand{\showtodos}{}  % Comment out to hide todos

@@

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\linenumbers
\begin{abstract}
% Say what problem is
  Capturing control flow is the Achilles heel of Haskell-based
  deeply embedded domain specific languages.
  Rather than use
  the builtin control flow mechanisms, artificial control flow combinators
  are used instead.
% Say why it is an interesting problem
  However, capturing traditional control flow in an deeply embedded domain specific language
  would support the writing of programs in a natural style by allowing the programmer to use the
  constructs that are already builtin to the base language, such as pattern
  matching and recursion.
% Say what our solution achieves
  In this paper, we expand the capabilities of
  Haskell-based deep embeddings with a compiler extension
  for reifying conditionals and pattern matching.
% Here is why the world will be a better place
  With this new support, the Haskell that we capture for expressing
  deeply embedded domain specific languages can be cleaner, Haskell-idiomatic,
  and more declarative in nature.

@@

\begin{lstlisting}[deletekeywords={Ord}]
x :: Either Char Int
x = Left 'a'

example :: Int
example =
  externalize (case x of Left c -> ord c; Right i -> i)

@@

@@

\begin{todo}
Should we have a diagram somewhere in the paper to show how these pieces all fit together?
\end{todo}
% \section{Citations and Bibliographies}

@@

% \appendix
% \section{Additional Code for Standard Haskell Semantics Implementation}
% \begin{lstlisting}[basicstyle=\tiny]
% absEnv :: forall t. Env -> E t -> t
% absEnv env (CaseExp x f) =
%   sumMatchAbs f (absEnv env x)

% absEnv env (TailRec f) = \ x ->
%   case absEnv env f x of
%     Step x' -> absEnv env (TailRec f) x'
%     Done r  -> r

% absEnv env (Var v) =
%   case envLookup env v of
%     Just x -> absEnv env
%     Nothing ->
%       error
%         ("No binding for name "
%           ++ show v)

% absEnv env (Lam (name :: Name a)
%                 (body :: E b)) =
%   \ (arg :: a) ->
%     let go :: forall x. E x -> E x
%         go expr@(Var name2 :: E a') =
%           case namesEq name name2 of
%             Just Refl -> rep arg
%             _ ->
%               case envLookup env name2 of
%                 Just v  -> v
%                 Nothing -> expr

%         -- [... traverse rest of expression
%         --      in go ...]
%     in
%     absEnv (extendEnv env
%                       (name :=> rep arg))
%            (go body)

% absEnv env (LeftExp x) =
%   Left (absEnv env x)

% absEnv env (RightExp y) =
%   Right (absEnv env y)

% absEnv env (PairExp x y) =
%   (absEnv env x, absEnv env y)

% absEnv env (ConstructRep x) =
%   unrep' (absEnv env x)

% ...
% \end{lstlisting}

% % \vfill\break

% \section{A small EDSL for C code generation used internally by the C backend}
% \begin{lstlisting}[basicstyle=\tiny]
% type CCode = String
% type CName = String

% stmt :: CCode -> CCode
% stmt = (<> ";")

% cCall :: CName -> [CCode] -> CCode
% cCall fnName args =
%   fnName
%   <> "(" <> intercalate ", " args <> ")"

% derefPtr :: CCode -> CCode
% derefPtr x = "*(" <> x <> ")"

% cCast :: CCode -> CCode -> CCode
% cCast toType x =
%   "((" <> toType <> ")" <> x <> ")"

% cIndex :: CCode -> Int -> CCode
% cIndex x i =
%   "(" <> x <> ")[" <> show i <> "]"

% (!) :: CCode -> Int -> CCode
% (!) = cIndex

% (=:) :: CCode -> CCode -> CCode
% x =: y = stmt (x <> " = " <> y)

% cDecl :: CName -> CName -> CCode
% cDecl cType v = stmt (cType <> " " <> v)

% (#) :: CCode -> CName -> CCode
% x # fieldName = x <> "." <> fieldName

% macroDef
%   :: CName -> [CName] -> [CCode]
%      -> CCode
% macroDef name args theLines =
%   unlines
%     [ "#define " <> name
%               <> "(" <>
%                 intercalate ", " args
%               <> ")\\"
%     , "  do {\\"
%     , unlines
%         . map ("    "++)
%         . map (++"\\")
%         $ theLines
%     , "  } while (0);"
%     ]
% \end{lstlisting}
